Income's Impact on Food Spending 

### 2024-04-18 paper/paper.qmd
updated the title

---
title: "You have more money in your wallet. This does not mean you will spend extra on food"
subtitle: "Examining the relationship between disposable income and food expenditure for American consumers from 2007-2022"
author: Ahnaf Alam
thanks: "Code and data are available at https://github.com/AhnafAlam1/income_and_expenditure"
date: today
date-format: long
classoption: abstract
abstract: "We examined the relationship between changes in food spending and increases in disposable income. Utilizing quarterly country-level data from the Federal Reserve Economic Data (FRED) spanning a 15-year period from 2007 to 2022, we employed both a simple linear regression model and a multiple regression model. Our analysis revealed a economically insignificant increase of $20 million in food spending for every billion-dollar rise in disposable income. From a policy perspective, our findings indicate that solely targeting disposable income is unlikely to substantially impact food spending behaviors. Therefore, policies designed to boost consumer spending should explore alternative approaches."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
margin-left: 25.4mm
margin-right: 25.4mm
---

@@

library(tidyverse)
library(ggplot2)
library(rstanarm)
library(kableExtra)
library(viridis)
library(testthat)
library(fredr)

@@

We utilize two linear models in our analysis: simple linear regression and multiple regression. Simple linear regression enables us to easily quantify the linear association between food expenditure and disposable income, as illustrated by @fig-3. However, this model has limitations. Most notably, it does not account for confounding variables, and it requires us to make unrealistic assumptions about the independence of the error term ($\epsilon$) and the predictor variable, income. This assumption implies that food expenditure and income are entirely independent of each other, which is problematic. Factors like economic downturns, income levels, and other variables can correlate with our model variables.

To address these issues, we employ a multiple linear regression model. This model helps us control for potential confounding variables, such as expenditure on non-durable goods, which may influence consumer expenditure on food. Using multiple linear regression allows us to better isolate the effect of income on food expenditure.

>>>>

---
title: "Income's Impact on Food Spending "
subtitle: "Examining the relationship between disposable income and food expenditure for American consumers from 2007-2022"
author: Ahnaf Alam
thanks: "Code and data are available at https://github.com/AhnafAlam1/income_and_expenditure"
date: today
date-format: long
classoption: abstract
abstract: "We examined the relationship between changes in food spending and increases in disposable income. Utilizing quarterly country-level data from the Federal Reserve Economic Data (FRED) spanning a 15-year period from 2007 to 2022, we employed both a simple linear regression model and a multiple regression model. Our analysis revealed a economically insignificant increase of $20 million in food spending for every billion-dollar rise in disposable income. From a policy perspective, our findings indicate that solely targeting disposable income is unlikely to substantially impact food spending behaviors. Therefore, policies designed to boost consumer spending should explore alternative approaches."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
margin-left: 25.4mm
margin-right: 25.4mm
---

@@

library(tidyverse)
library(ggplot2)
library(rstanarm)
library(kableExtra)
library(viridis)
library(testthat)
library(fredr)
library(DiagrammeR)

@@

We employ two linear models in our analysis: simple linear regression and multiple regression. Simple linear regression provides a straightforward way to quantify the linear relationship between food expenditure and disposable income, as depicted in @fig-3. However, this model has limitations. Specifically, it does not consider confounding variables and relies on unrealistic assumptions about the independence between the error term ($\epsilon$) and the predictor variable, income. This assumption suggests that food expenditure and income are completely independent, which is not the case. Variables like economic downturns and income levels can correlate with our model variables.

To overcome these limitations, we use a multiple linear regression model. This approach allows us to control for potential confounding variables, such as expenditure on non-durable goods, which can influence consumer spending on food. For instance, when disposable income rises, people might choose to prioritize spending on non-durable goods over food, leading to a decrease in food expenditure. This means that non-durable spending directly affects income and indirectly impacts food expenditure—a relationship that simple linear regression fails to capture.

We hypothesize that durable goods, non-durable goods, healthcare spending, and services spending are key confounding variables of interest because they represent a significant portion of consumer spending. By employing multiple linear regression, we can account for these variables and more accurately isolate the effect of income on food expenditure.
